```{r setup, include=FALSE, echo=FALSE}
### Set knitr global options
knitr::opts_chunk$set(warning=FALSE, message=FALSE)

### Establish color schemes and themes needed

theme_sara_90 <- function() {
  theme_bw(base_size=18)+
    theme(axis.text.x=element_text(angle=90,hjust = 1,vjust = 0.5),
          axis.text=element_text(color="black"),
          panel.background=element_rect(color="black"),
          strip.text = element_text(size=12),
          strip.background = element_rect(fill="white"))
}

### Cluster options
cluster_option_readable = dplyr::case_when(params$dex_cluster_algorithm == 1 ~ "Original Louvain algorithm",
                                           params$dex_cluster_algorithm == 2 ~ "Louvain with multilevel refinement",
                                           params$dex_cluster_algorithm == 3 ~ "SLM algorithm",
                                           params$dex_cluster_algorithm == 4 ~ "Leiden algorithm")

### If more colors are needed, use the following-------------------------------
#library(randomcoloR)
#n <- 3
#palette <- distinctColorPalette(n)
#pie(rep(1, n), col=palette)
```

```{r, echo = FALSE, include = FALSE}
if(!exists("seurats")){
  seurats = readRDS(params$anno_save_name)
}

### Chang identity to cell type
for(i in 1:length(seurats)){
    Idents(seurats[[i]])<-seurats[[i]]@meta.data[,params$dex_ident]
}
```

## 3.1 - PCA analyses {.tabset .tabset-pils}

Perform linear dimensional reduction via PCA.
We use different methods to run diagnostics on the Principal Components (PCs).

### Print the Principal Components (PCs) 

```{r, tidy='styler', tidy.opts=list(strict=FALSE)}
### Perform linear dimensional reduction via PCA
seurats <- lapply(X = seurats, FUN = function(x) {
    x <- RunPCA(x)
})
```

```{r, echo=FALSE, tidy='styler', tidy.opts=list(strict=FALSE)}
# print(seurats[["pca"]], dims = 1:15, nfeatures = 5)
aux<-list()
features_pca<-list()
for(i in 1:length(seurats)) {
aux[[i]] = seurats[[1]][["pca"]][,1:15] |>
  as.data.frame() |>
  tibble::rownames_to_column(var = "gene") |>
  tidyr::pivot_longer(cols = !gene, names_to = "PCs", values_to = "values") |>
  dplyr::group_by(PCs) |>
  dplyr::arrange(values, .by_group = TRUE) |>
  dplyr::top_n(-5)

features_pca[[i]] = seurats[[i]] [["pca"]][,1:15] |>
  as.data.frame() |>
  tibble::rownames_to_column(var = "gene") |>
  tidyr::pivot_longer(cols = !gene, names_to = "PCs", values_to = "values") |>
  dplyr::group_by(PCs) |>
  dplyr::arrange(desc(values), .by_group = TRUE) |>
  dplyr::top_n(5) |>
  dplyr::bind_rows(aux[[i]]) |>
  dplyr::mutate(direction = dplyr::if_else(values < 0, "Negative", "Positive")) |>
  tidyr::pivot_wider(id_cols = PCs, names_from = "direction", values_from = "gene")

features_pca[[i]] |>
  dplyr::mutate(PCs = factor(gsub("_", " ", PCs), levels = paste0("PC ", 1:15))) |>
  dplyr::arrange(PCs) |>
  tibble::column_to_rownames(var = "PCs") 
}

### Print it out
features_pca[[i]] |> render_dt()
```

<br>
<br>

### PCA loadings {.tabset .tabset-pils}

This plot is to identify genes contributing to desired PCAs

```{r, tidy='styler', results="asis", tidy.opts=list(strict=FALSE), fig.cap="Identify genes contributing to desired PCAs"}
### Generate a list of plots based on the above information
plots<-list()
for (i in 1:length(seurats)) {
    p1 <- eval(substitute(
        
        VizDimLoadings(seurats[[i]],
                       dims = 1:4,
                       reduction = "pca")

        ,list(i = i)))
    plots[[i]] <- p1  # add each plot into plot list
}


#### Plot it out
for (i in 1:length(plots)){
print(glue::glue("#### {as.vector(names(seurats[i]))}\n"))
  
title_plots <- plots[[i]]
print(title_plots)
cat("\n \n \n")
}
```

<br>
<br>

### PCA by cell types {.tabset .tabset-pils}

```{r, results="asis", tidy='styler', tidy.opts=list(strict=FALSE)}
plots<-list()
for (i in 1:length(seurats)) {
    p1 <- eval(substitute(
        
        DimPlot(seurats[[i]], reduction = "pca",
                cols = cell.cols,
                pt.size = 0.5) +
            ggtitle("PCA of cell types defined by SingleR") +
            themes$QB_theme()+
            ylab("PC 2")+
            xlab("PC 1")+
            theme(axis.title.x = element_text())
        
        ,list(i = i)))
    plots[[i]] <- p1  # add each plot into plot list
}

#### Plot it out
for (i in 1:length(plots)){
print(glue::glue("#### {as.vector(names(seurats[i]))}\n"))
  
title_plots <- plots[[i]]
print(title_plots)
cat("\n \n \n")
}
```

<br>
<br>

### PCA by features {.tabset .tabset-pils}

```{r, results="asis", tidy='styler', tidy.opts=list(strict=FALSE)}
plots<-list()
for (i in 1:length(seurats)) {
    p1 <- eval(substitute(
        
        FeaturePlot(seurats[[i]],
                    reduction='pca',
                    features=c("nFeature_RNA",  "percent.mt"),
                    raster=FALSE)
                ,list(i = i)))
    plots[[i]] <- p1  # add each plot into plot list
}

#### Plot it out
for (i in 1:length(plots)){
print(glue::glue("#### {as.vector(names(seurats[i]))}\n"))
  
title_plots <- plots[[i]]
print(title_plots)
cat("\n \n \n")
}

```

<br>
<br>

### PCA by phase {.tabset .tabset-pils}

```{r,tidy='styler', results="asis",tidy.opts=list(strict=FALSE), eval=params$qcs_cc}
for(i in 1:length(seurats)){
    Idents(seurats[[i]])<-seurats[[i]]@meta.data$Phase
}

plots<-list()
for (i in 1:length(seurats)) {
    p1 <- eval(substitute(
        
        DimPlot(seurats[[i]], reduction = "pca",
                cols = phase.colors,
                pt.size = 0.5) +
            ggtitle("PCA of cells colored by cell cycle phase") +
            themes$QB_theme()+
            ylab("PC 2")+
            xlab("PC 1")+
            theme(axis.title.y = element_text(),
                  axis.title.x = element_text())
        
        ,list(i = i)))
    plots[[i]] <- p1  # add each plot into plot list
}

#### Plot it out
for (i in 1:length(plots)){
print(glue::glue("#### {as.vector(names(seurats[i]))}\n"))
  
title_plots <- plots[[i]]
print(title_plots)
cat("\n \n \n")
}

### Return identities
for(i in 1:length(seurats)){
    Idents(seurats[[i]])<-seurats[[i]]@meta.data[,params$dex_ident]
}
```

### Dimension heatmap {.tabset .tabset-pils}

This method allows for easy exploration of the primary sources of heterogeneity in a dataset, 
and can be useful when trying to decide which PCs to include for further downstream analyses.

```{r, results="asis",tidy='styler'}
#### Plot it out
for (i in 1:length(seurats)){
print(glue::glue("#### {as.vector(names(seurats[i]))}\n"))
title_plots <- print(DimHeatmap(seurats[[i]], dims = 1:15, cells = 500, balanced = TRUE))
cat("\n \n \n")
}
```

<br>
<br>

### Elbow plot {.tabset .tabset-pils}

We look for an ‘elbow’ in the plot (usually around PC6-7).
However, even PCs up to usually 12 still maintain a high standard deviation.

This is supported by the heatmaps, suggesting that the majority of 
true signal is captured in the first 12 PCs.

```{r, results="asis", fig.cap="Identify most influential PCs", tidy='styler', tidy.opts=list(strict=FALSE)}
### Make some more lists
pcas<-list()
eigValues<-list()
varExplained<-list()
PCs<-list()
ndims<-list()

### Determine which PCs explain 80% of the variance
for(i in 1:length(seurats)){
    
    pcas[[i]] = seurats[[i]]@reductions$pca
    eigValues[[i]] = (pcas[[i]]@stdev)^2 ## EigenValues
    varExplained[[i]]=eigValues[[i]]/sum(eigValues[[i]])
    PCs[[i]]<-sum(ifelse(cumsum(varExplained[[i]])<=params$dex_variance_explained, 1, 0))
    if (PCs[[i]] < 5){PCs[[i]] = 5}
    ndims[[i]] = ifelse(PCs[[i]] >= 20, PCs[[i]] + 1, 30)
    
}

### Make some more plots
plots<-list()
for (i in 1:length(seurats)) {
    p1 <- eval(substitute(
        
        ElbowPlot(seurats[[i]], ndims = ndims[[i]]) +
            ggplot2::geom_vline(xintercept = PCs[[i]], linetype='dashed', color='blue')

        ,list(i = i)))
    plots[[i]] <- p1  # add each plot into plot list
}

#### Plot it out
for (i in 1:length(plots)){
print(glue::glue("#### {as.vector(names(seurats[i]))}\n"))
  
title_plots <- plots[[i]]
print(title_plots)
cat("\n \n \n")
}
```

<br>
<br>

We use **PCs** that capture the **`r params$dex_variance_explained*100`% of the variance**.
This is a good rule of thumb for most datasets, and is a good starting point for most analyses.

## 3.2 - Confounding variables {.tabset .tabset-pils}

### Categorical and numerical vars vs. PCs {.tabset .tabset-pils}

Variables that are associated with the top 10 PCs. These variables may be confounding variables and worth considering in downstream analyses. 

```{r,confounding, results="asis", fig.cap="Variables that are associated with the top 10 PCs. These variables may be confounding variables and worth considering in downstream analyses. ", tidy='styler', tidy.opts=list(strict=FALSE), eval=params$dex_confounding}
### Calculate PCs
mats<-list()
pca.list<-list()

### Calculate PC statistics (again)
for(i in 1:length(seurats)){
    mats[[i]]<-as.data.frame(LayerData(seurats[[i]], assay="RNA", layer="counts"))
    pca.list[[i]]<-list(All=FactoMineR::PCA(t(mats[[i]]), ncp=10, graph=FALSE, scale=FALSE))
}    

### Calculate covariates
anova.results<-list()
glm.results<-list()
results<-list()
df.melt<-list()
for(i in 1:length(pca.list)) {
    ### [1] Run ANOVA of categorical variables
    anova.results[[i]] <- anova_PCA(obj_name="All",
                               pca.df=pca.list[[i]]$All,
                               metadata=seurats[[i]]@meta.data,
                               merge_ID="Cell",
                               cat_var=params$dex_categorical_vars,
                           numPCs=10)
    rownames(anova.results[[i]])<-anova.results[[i]]$Factor

    ### [2] Run GLM for continuous variables
    glm.results[[i]] <- glm_PCA(obj_name="All",
                                pca.list=pca.list[[i]]$All,
                                metadata=seurats[[i]]@meta.data,
                                merge_ID="Cell",
                                num_var=params$dex_numerical_vars,
                                numPCs=10)
    rownames(glm.results[[i]])<-glm.results[[i]]$Factor

    ### [3] Melt results for plotting
    results[[i]] <- rbind(anova.results[[i]], glm.results[[i]])
    df.melt[[i]] <- melt(results[[i]])
    df.melt[[i]] <- cbind(df.melt[[i]], PC=as.numeric(gsub("Dim|_adj_pvalue", "", df.melt[[i]]$variable)))
    df.melt[[i]]$ajp <- ifelse(df.melt[[i]]$value <= 0.001, "<=0.001", df.melt[[i]]$value)
    df.melt[[i]]$ajp <- ifelse(df.melt[[i]]$ajp <= 0.01 &  df.melt[[i]]$ajp > 0.001, "<=0.01",  df.melt[[i]]$ajp)
    df.melt[[i]]$ajp <- ifelse(df.melt[[i]]$ajp <= 0.05 &  df.melt[[i]]$ajp > 0.01, "<=0.05",  df.melt[[i]]$ajp)
    df.melt[[i]]$ajp <- ifelse(df.melt[[i]]$ajp > 0.05, "NS",  df.melt[[i]]$ajp)
}

### Plot
plots<-list()
for (i in 1:length(seurats)) {
    p1 <- eval(substitute(

        ggplot(df.melt[[i]], aes(x=PC, y=Factor)) +
            geom_tile(aes(fill=ajp),color="white") +
            scale_fill_manual(values=c("<=0.001"="#4c1d4bff",
                                       "<=0.01"="#bd1655ff",
                                       "<=0.05"="#f47f58ff",
                                       "NS"="#FAEBddff"),
                              name="FDR adjusted p-value") +
            labs(title="Covariates vs. PCs",
                 subtitle=paste0("Data is from: ", names(seurats[i]))) +
            themes$QB_theme()+
            ylab("")
    

                ,list(i = i)))
    plots[[i]] <- p1  # add each plot into plot list
    }

#### Plot it out
for (i in 1:length(plots)){
print(glue::glue("#### {as.vector(names(seurats[i]))}\n"))

title_plots <- plots[[i]]
print(title_plots)
cat("\n \n \n")
}
```

<br>
<br>

## 3.3 - uMAP and tSNE visualisations {.tabset .tabset-pils}

***
**NOTE** 

It is encouraged users to repeat downstream analyses with a different number of 
PCs (10, 15, or even 50!). As the results often do not differ dramatically.
It is also advised users to err on the higher side when choosing this parameter. 
Because downstream analyses with only 5 PCs do significantly and adversely affect results.

***

Regarding clustering methods, `SeuratV3` embeds cells in a graph structure - for example a 
K-nearest neighbor (KNN) graph, with edges drawn between cells with similar feature 
expression patterns, and then attempts to partition this graph into highly interconnected 
_‘quasi-cliques’_ or _‘communities’_.

We first construct a KNN graph based on the euclidean distance in PCA space, and 
refine the edge weights between any two cells based on the shared overlap in their 
local neighborhoods (Jaccard similarity). 
This step is performed using the `FindNeighbors()`

To cluster the cells, we next apply modularity optimization techniques 
implemented in the function `FindClusters()` from the _Seurat package_.

Available options are the Louvain algorithm (default) or SLM [Blondel et al. 2008](https://iopscience.iop.org/article/10.1088/1742-5468/2008/10/P10008), 
that iteratively group cells together, with the goal of optimizing the 
standard modularity function.
Optimal resolution for 3K cells often ranges from 0.4 to 1.2 and increases the larger
the dataset.

The option selected for this analysis is **`r cluster_option_readable`**.

### UMAP colored by cell type {.tabset .tabset-pils}

```{r, tidy='styler', results="asis",tidy.opts=list(strict=FALSE), comment= ""}
### Calculate UMAP, tSCNE, and identify clusters
for(i in 1:length(seurats)){
    seurats[[i]] <- FindNeighbors(seurats[[i]], dims = 1:PCs[[i]])
    seurats[[i]] <- FindClusters(seurats[[i]], 
                            algorithm = params$dex_cluster_algorithm)
    seurats[[i]] <- RunUMAP(seurats[[i]], dims = 1:PCs[[i]])
    seurats[[i]] <- RunTSNE(seurats[[i]], dims = 1:PCs[[i]], check_duplicates=FALSE)
}


### Make some more plots
plots<-list()
for (i in 1:length(seurats)) {
    p1 <- eval(substitute(
        
        DimPlot(seurats[[i]], reduction = "umap",
                group.by = "SingleR2",
                cols = cell.cols,
                #        shuffle = TRUE,
                #        repel=TRUE,
                #        label=TRUE, label.size = 2.5,label.box = TRUE,
                pt.size = 0.5) +
            ggtitle("UMAP of cell types defined by SingleR") +
            themes$QB_theme()+
            ylab("UMAP 2")+
            xlab("UMAP 1")+
            theme(axis.title.y = element_text(),
                  axis.title.x = element_text())
        
        ,list(i = i)))
    plots[[i]] <- p1  # add each plot into plot list
    }

#### Plot it out
for (i in 1:length(plots)){
print(glue::glue("#### {as.vector(names(seurats[i]))}\n"))
  
title_plots <- plots[[i]]
print(title_plots)
cat("\n \n \n")
}
```

<br>
<br>

### UMAP colored by cell cycle phase {.tabset .tabset-pils}

```{r,  tidy='styler', results="asis", tidy.opts=list(strict=FALSE), comment= "", eval=params$qcs_cc}
### Color by cell cycle phase
### Make some more plots
plots<-list()
for (i in 1:length(seurats)) {
    p1 <- eval(substitute(
        
        DimPlot(seurats[[i]], reduction = "umap",
        group.by = "Phase",
        cols = phase.colors,
#        shuffle = TRUE,
#        repel=TRUE,
#        label=TRUE, label.size = 2.5,label.box = TRUE,
        pt.size = 0.5) +
  ggtitle("UMAP of cell types colored by cell cycle phase") +
  themes$QB_theme()+
  ylab("UMAP 2")+
  xlab("UMAP 1")+
  theme(axis.title.y = element_text(),
        axis.title.x = element_text())

        ,list(i = i)))
    plots[[i]] <- p1  # add each plot into plot list
    }

#### Plot it out
for (i in 1:length(plots)){
print(glue::glue("#### {as.vector(names(seurats[i]))}\n"))
  
title_plots <- plots[[i]]
print(title_plots)
cat("\n \n \n")
}
```

## 3.4 Differential expression per cluster

Seurat can help you find markers that define clusters via differential expression. 
By default, it identifies positive and negative markers of a single cluster (specified 
in `ident.1`), compared to all other cells. 
`FindAllMarkers()` automates this process for all clusters, 
but you can also test groups of clusters vs. each other, or against all cells.

The `min.pct` argument requires a feature to be detected at a minimum percentage in either 
of the two groups of cells, 
and the `thresh.test` argument requires a feature to be 
differentially expressed (on average) by some amount between the two groups. 
You can set both of these to 0, 
but with a dramatic increase in time - since this will test a large 
number of features that are unlikely to be highly discriminatory.
Currently, the values for `min.pct` and `thresh.test` are
`r params$dex_min_pct` and `r params$dex_logfc_threshold`, respectivelly.

As another option to speed up these computations, 
`max.cells.per.ident` can be set. 
This will downsample each identity class to have no more cells than whatever this is set to. 
While there is generally going to be a loss in power, 
the speed increases can be significant and the 
most highly differentially expressed features will likely still rise to the top.

```{r, tidy='styler', tidy.opts=list(strict=FALSE),eval=params$dex_run_dex}
if (file.exists(params$dex_mat_save_name)){
  DE_seurats <- readRDS(params$dex_mat_save_name)
  text<-paste0("DEX data source: ","DE file read successfully.")
} else {
    ### Find markers for every cluster compared to all remaining cells
    DE_seurats<-list()
    for(i in 1:length(seurats)){
    Idents(seurats[[i]])<-seurats[[i]]@meta.data[,params$dex_ident]
    DE_seurats[[i]] <- FindAllMarkers(seurats[[i]], 
                            only.pos = params$dex_only_positive, 
                             min.pct = params$dex_min_pct, 
                             logfc.threshold = params$dex_logfc_threshold)
    DE_seurats[[i]]$params_group<-names(seurats[i])
    }
    
    ### Write to file
    write_rds(DE_seurats, file=params$dex_mat_save_name)
    text<-paste0("DEX data source: ","DE file created and data written successfully.")
}
```


### DEGs heatmap {.tabset .tabset-pils}

```{r, results="asis", fig.height=14, fig.width=16, fig.cap="Heatmap of top 5 Differentially Expressed Genes", tidy='styler', tidy.opts=list(strict=FALSE),eval=params$dex_run_dex}
### Make a list of a list. These are the items necessary to make a heatmap
heatmap_stuff<-list()
for(i in 1:length(DE_seurats)) {
    heatmap_stuff[[i]]<-LetsMakeAHeatmap(dex_data=DE_seurats[[i]], 
                         pval = 0.05, 
                         top_n = 5, 
                         log2FC = 0.0,
                         sample_size = 250, 
                         seurat = seurats[[i]], 
                         variable = "SingleR2")
}

### Make the annotation objects
has<-list()
ras<-list()
morecols<-list()
has<-for(i in 1:length(heatmap_stuff)) {
    
    ### 1) Column annotations
    has[[i]] <- HeatmapAnnotation(`TEMPUS cell type`=heatmap_stuff[[i]]$seurat_sample$SingleR2,
                        col=list(`Group`=cell.cols
                                 ),
                        annotation_name_side = "left",
                        border=TRUE,
                        simple_anno_size = unit(0.35, "cm"))
    
    ### 2) Row annotations
    ras[[i]]<- rowAnnotation(`Fold-change`=anno_barplot(heatmap_stuff[[i]]$top$avg_log2FC, 
                                               add_numbers=F),
                           `Group source` = heatmap_stuff[[i]]$top$cluster,
                           `Dup.`=heatmap_stuff[[i]]$top$Duplicated,
                           col=list(`Group source`=cell.cols,
                                    `Dup.`=c("Yes"="yellow",
                                             "No"="purple")
                                    ),
                    gp = gpar(col = "black"),
                    annotation_name_rot=90,
                    show_legend = c(`Sample source`=FALSE)
                        )
    
    ### 3) Color scheme 3 for expression levels
    morecols[[i]] <- colorRamp2( c(0,max(as.matrix(heatmap_stuff[[i]]$mat)),
                          max(heatmap_stuff[[1]]$mat)), c("white","orangered2","black") )
    }


### Plot the heatmap for tumor tissue
maps<-list()
for(i in 1:length(heatmap_stuff)) {
    
    maps[[i]]<-Heatmap(heatmap_stuff[[i]]$mat, 
        name="Expression", 
        col=morecols[[i]],         
        border = T, 
        border_gp = gpar(col = "darkgrey", lty = 1),
        
        ### Column stuff
        top_annotation = has,
        show_column_names = F,
        column_split = factor(heatmap_stuff[[i]]$seurat_sample$SingleR2,
                             levels=heatmap_stuff[[i]]$lvls$Var1),

        cluster_column_slices = F,
        column_gap = unit(1.0,"mm"),
        column_title_side = "top",
        column_title=paste0(names(seurats[i]), " cells only"),

        ### Row stuff
        left_annotation = ras[[i]],
        show_row_names = F,
        row_split=factor(heatmap_stuff[[i]]$top$cluster,
                         levels=heatmap_stuff[[i]]$lvls$Var1),
        cluster_row_slices =  F,
        row_gap = unit(1.0, "mm"),
        row_title = "Top DEGs per celltype",
        row_names_gp = gpar(fontsize = 7)
        )
    
}

### Print the heatmaps
for (i in 1:length(seurats)){
print(glue::glue("#### {as.vector(names(seurats[i]))}\n"))
title_plots <- print(maps[[i]])
cat("\n \n \n")
}

```

## 3.5 - Write seurat to file
The annotation is done, the file is save on **_`r params$dex_seurat_save_name`_**.

```{r, include=FALSE, eval=params$dex_write_dex_seurat}
### All the QC is done, so now we may write to file
write_rds(seurats, file=params$dex_seurat_save_name)
```
