```{r setup, include=FALSE, echo=FALSE}
### Set knitr global options
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)

### Load libraries
library(data.table)
library(ComplexHeatmap)
library(annotate)
library(SingleR)
library(celldex)
library(Seurat)
library(scales)
library(RColorBrewer)
library(ggplot2)
library(plyr)
library(kableExtra)
library(knitr)
library(forcats)
library(readxl)
library(dplyr)
library(ggpubr)
library(circlize)
library(readr)

### Establish color schemes and themes needed
theme_sara <- function() {
  theme_bw(base_size=14)+
    theme(axis.text=element_text(color="black"),
          panel.background=element_rect(color="black"),
          strip.text = element_text(size=12),
          strip.background = element_rect(fill="white"))
}

theme_sara_90 <- function() {
  theme_bw(base_size=18)+
    theme(axis.text.x=element_text(angle=90,hjust = 1,vjust = 0.5),
          axis.text=element_text(color="black"),
          panel.background=element_rect(color="black"),
          strip.text = element_text(size=12),
          strip.background = element_rect(fill="white"))
}

ct.colors <- c("T_cells"="chartreuse4",
               "B_cell"="darkorange",
               "NK_cell"="red",
               "Epithelial_cells"="firebrick", 
               "Endothelial_cells"="hotpink", 
               "Fibroblasts"="royalblue", 
               "Macrophage"="chocolate4", 
               "CMP"="navy",
               "Others"="darkorchid",
               "DC"="cyan4",
               "Monocyte"="bisque3")

cols <- colorRamp2( c(0.0,0.5,8.25), c("white","orangered2","black"))

phase.colors <- c("G1"="#BF6ECA",
               "G2M"="#ACDE7C",
               "S"="#BEBEBB")

cluster_option_readable = dplyr::case_when(params$dex_cluster_algorithm == 1 ~ "Original Louvain algorithm",
                                           params$dex_cluster_algorithm == 2 ~ "Louvain with multilevel refinement",
                                           params$dex_cluster_algorithm == 3 ~ "SLM algorithm",
                                           params$dex_cluster_algorithm == 4 ~ "Leiden algorithm")

### If more colors are needed, use the following-------------------------------
#library(randomcoloR)
#n <- 3
#palette <- distinctColorPalette(n)
#pie(rep(1, n), col=palette)
```

```{r, echo = FALSE, include = FALSE}
if(!exists("seurat_analysed")){
  seurat_analysed = readRDS(params$anno_save_file)
}

### Chang identity to cell type
Idents(seurat_analysed)<-seurat_analysed@meta.data[,params$dex_ident_cell_type]
# seurat_analysed<-subset(seurat_analysed, downsample=1000)

```

## 3.1 - PCA analyses

Perform linear dimensional reduction via PCA.
We use different methods to run diagnostics on the Principal Components (PCs).

### Method 1 - Print the Principal Components (PCs) 

```{r, tidy='styler', tidy.opts=list(strict=FALSE)}
### Perform linear dimensional reduction via PCA
seurat_analysed <- RunPCA(seurat_analysed)
```

```{r, echo=FALSE, tidy='styler', tidy.opts=list(strict=FALSE)}
# print(seurat_analysed[["pca"]], dims = 1:15, nfeatures = 5)

aux = seurat_analysed[["pca"]][,1:15] %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "gene") %>%
  tidyr::pivot_longer(cols = !gene, names_to = "PCs", values_to = "values") %>%
  dplyr::group_by(PCs) %>%
  dplyr::arrange(values, .by_group = TRUE) %>%
  dplyr::top_n(-5)

features_pca = seurat_analysed[["pca"]][,1:15] %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "gene") %>%
  tidyr::pivot_longer(cols = !gene, names_to = "PCs", values_to = "values") %>%
  dplyr::group_by(PCs) %>%
  dplyr::arrange(desc(values), .by_group = TRUE) %>%
  dplyr::top_n(5) %>%
  dplyr::bind_rows(., aux) %>%
  dplyr::mutate(direction = dplyr::if_else(values < 0, "Negative", "Positive")) %>%
  tidyr::pivot_wider(id_cols = PCs, names_from = "direction", values_from = "gene")

features_pca %>%
  dplyr::mutate(PCs = factor(gsub("_", " ", PCs), levels = paste0("PC ", 1:15))) %>%
  dplyr::arrange(PCs) %>%
  tibble::column_to_rownames(var = "PCs") %>%
  knitr::kable(.,
               caption = "Top 5 features in the first 15 Principal Components") %>%
  kableExtra::column_spec(1, bold = T)
```

### Method 2 - VizDimLoadings
This plot is to identify genes contributing to desired PCAs

```{r, tidy='styler', tidy.opts=list(strict=FALSE), fig.cap="Identify genes contributing to desired PCAs"}
VizDimLoadings(seurat_analysed, 
               dims = 1:4, 
               reduction = "pca")
```

### Method 3 - Dimplot
```{r, tidy='styler', tidy.opts=list(strict=FALSE)}
DimPlot(seurat_analysed, reduction = "pca",
        cols = ct.colors,
        pt.size = 0.5) +
  ggtitle("PCA of cell types defined by SingleR") +
  theme_sara()+
  ylab("PC 2")+
  xlab("PC 1")+
  theme(axis.title.x = element_text())

FeaturePlot(seurat_analysed,  
            reduction='pca', 
            features=c("nFeature_RNA",  "percent.mt"))

Idents(seurat_analysed)<-seurat_analysed@meta.data$Phase
DimPlot(seurat_analysed, reduction = "pca",
        cols = phase.colors,
        pt.size = 0.5) +
  ggtitle("PCA of cells colored by cell cycle phase") +
  theme_sara()+
  ylab("PC 2")+
  xlab("PC 1")+
  theme(axis.title.y = element_text(),
        axis.title.x = element_text())
Idents(seurat_analysed)<-seurat_analysed@meta.data$SingleR2
```

### Method 4 - DimHeatmapt 

This method allows for easy exploration of the primary sources of heterogeneity in a dataset, 
and can be useful when trying to decide which PCs to include for further downstream analyses.

```{r, tidy='styler', tidy.opts=list(strict=FALSE)}
DimHeatmap(seurat_analysed, dims = 1:15, cells = 500, balanced = TRUE)
```

### Method 5 - Elbow plot

We look for an ‘elbow’ in the plot (usually around PC6-7).
However, even PCs up to usually 12 still maintain a high standard deviation.

This is supported by the heatmaps, suggesting that the majority of 
true signal is captured in the first 12 PCs.

```{r, fig.cap="Identify most influential PCs", tidy='styler', tidy.opts=list(strict=FALSE)}
pca = seurat_analysed@reductions$pca
eigValues = (pca@stdev)^2 ## EigenValues
varExplained=eigValues/sum(eigValues)

PCs<-sum(ifelse(cumsum(varExplained)<=params$dex_variance_explained, 1, 0))

if (PCs < 5){PCs = 5}

ndims = ifelse(PCs >= 20, PCs + 1, 30)

ElbowPlot(seurat_analysed, ndims = ndims) +
  ggplot2::geom_vline(xintercept = PCs, linetype='dashed', color='blue')

```

We use **`r PCs` PCs** that capture the **`r params$dex_variance_explained*100`% of the variance**.

## 3.2 - uMAP and tSNE visualisations

***
**NOTE** 

It is encouraged users to repeat downstream analyses with a different number of 
PCs (10, 15, or even 50!). As the results often do not differ dramatically.
It is also advised users to err on the higher side when choosing this parameter. 
Because downstream analyses with only 5 PCs do significantly and adversely affect results.

***

Regarding clustering methods, `SeuratV3` embeds cells in a graph structure - for example a 
K-nearest neighbor (KNN) graph, with edges drawn between cells with similar feature 
expression patterns, and then attempts to partition this graph into highly interconnected 
_‘quasi-cliques’_ or _‘communities’_.

We first construct a KNN graph based on the euclidean distance in PCA space, and 
refine the edge weights between any two cells based on the shared overlap in their 
local neighborhoods (Jaccard similarity). 
This step is performed using the `FindNeighbors()`

To cluster the cells, we next apply modularity optimization techniques 
implemented in the function `FindClusters()` from the _Seurat package_.

Available options are the Louvain algorithm (default) or SLM [Blondel et al. 2008](https://iopscience.iop.org/article/10.1088/1742-5468/2008/10/P10008), 
that iteratively group cells together, with the goal of optimizing the 
standard modularity function.
Optimal resolution for 3K cells often ranges from 0.4 to 1.2 and increases the larger
the dataset.

The option selected for this analysis is **`r cluster_option_readable`**.

```{r, tidy='styler', tidy.opts=list(strict=FALSE), comment= ""}

### Identify neighbors
seurat_analysed <- FindNeighbors(seurat_analysed, dims = 1:PCs)

### Identify clusters
seurat_analysed <- FindClusters(seurat_analysed, 
                                algorithm = params$dex_cluster_algorithm)

### Run uMAP and tSNE
seurat_analysed <- RunUMAP(seurat_analysed, dims = 1:PCs)
seurat_analysed <- RunTSNE(seurat_analysed, dims = 1:PCs)

### Plot uMAP 
DimPlot(seurat_analysed, reduction = "umap",
        group.by = "SingleR2",
        cols = ct.colors,
#        shuffle = TRUE,
#        repel=TRUE,
#        label=TRUE, label.size = 2.5,label.box = TRUE,
        pt.size = 0.5) +
  ggtitle("UMAP of cell types defined by SingleR") +
  theme_sara()+
  ylab("UMAP 2")+
  xlab("UMAP 1")+
  theme(axis.title.y = element_text(),
        axis.title.x = element_text())

### Color by cell cycle phase
DimPlot(seurat_analysed, reduction = "umap",
        group.by = "Phase",
        cols = phase.colors,
#        shuffle = TRUE,
#        repel=TRUE,
#        label=TRUE, label.size = 2.5,label.box = TRUE,
        pt.size = 0.5) +
  ggtitle("UMAP of cell types colored by cell cycle phase") +
  theme_sara()+
  ylab("UMAP 2")+
  xlab("UMAP 1")+
  theme(axis.title.y = element_text(),
        axis.title.x = element_text())
```

## 3.3 Differential expression per cluster

Seurat can help you find markers that define clusters via differential expression. 
By default, it identifies positive and negative markers of a single cluster (specified 
in `ident.1`), compared to all other cells. 
`FindAllMarkers()` automates this process for all clusters, 
but you can also test groups of clusters vs. each other, or against all cells.

The `min.pct` argument requires a feature to be detected at a minimum percentage in either 
of the two groups of cells, 
and the `thresh.test` argument requires a feature to be 
differentially expressed (on average) by some amount between the two groups. 
You can set both of these to 0, 
but with a dramatic increase in time - since this will test a large 
number of features that are unlikely to be highly discriminatory.
Currently, the values for `min.pct` and `thresh.test` are
`r params$dex_min_pct` and `r params$dex_logfc_threshold`, respectivelly.

As another option to speed up these computations, 
`max.cells.per.ident` can be set. 
This will downsample each identity class to have no more cells than whatever this is set to. 
While there is generally going to be a loss in power, 
the speed increases can be significant and the 
most highly differentially expressed features will likely still rise to the top.

```{r, tidy='styler', tidy.opts=list(strict=FALSE)}
### Find markers for every cluster compared to all remaining cells
Idents(seurat_analysed)<-seurat_analysed@meta.data$SingleR2
DE_seurat_analysed <- FindAllMarkers(seurat_analysed, 
                            only.pos = params$dex_only_post, 
                             min.pct = params$dex_min_pct, 
                             logfc.threshold = params$dex_logfc_threshold)

sigs <- DE_seurat_analysed[which(DE_seurat_analysed$p_val_adj< params$dex_p_val_adj &
                           DE_seurat_analysed$pct.1>params$dex_min_pct &
                           DE_seurat_analysed$avg_log2FC>params$dex_logfc_threshold),]

top5 <- sigs %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC)

### List the top 5 highest expressed genes for each cell subtype
sigs %>%
    group_by(cluster) %>%
    slice_max(n = 5, order_by = avg_log2FC) %>%
    dplyr::select(gene, cluster, pct.1, pct.2, p_val, p_val_adj, avg_log2FC) %>%
    dplyr::mutate(p_val = as.character(signif(p_val, 2))) %>%
    dplyr::mutate(p_val_adj = as.character(signif(p_val_adj, 2))) %>%
    dplyr::rename(Pvalue = p_val, `Adjusted Pvalue` = p_val_adj, `log2FC avg` = avg_log2FC) %>%
    knitr::kable(.,
               caption = " Identify significantly DE genes. Selection of the 5 highest expressed genes for each cell subtype.",
               digits = 2) %>%
    kableExtra::scroll_box(width = "800px", height = "400px")

```


```{r, fig.height=10, fig.width=12, fig.cap="Heatmap of top 5 Differentially Expressed Genes", tidy='styler', tidy.opts=list(strict=FALSE)}
### Subset the top 5 genes
mat<-as.matrix(seurat_analysed@assays[["RNA"]]@data)
mat<-mat[top5$gene,]

### Create an object to establish column annotations
ha <- HeatmapAnnotation(Celltype=seurat_analysed@meta.data$SingleR2,
                        col=list(Celltype=ct.colors),
                        border=TRUE)

### Add row annotation colors
index<-match(top5$gene, DE_seurat_analysed$gene)
ra <- rowAnnotation(`Fold-change (FC)`=anno_barplot(top5$avg_log2FC, 
                                               add_numbers=TRUE),
                    `FC cell reference` = top5$cluster,
                    annotation_name_rot=90,
                    col=list(`FC cell reference`=ct.colors),
                    gp = gpar(col = "black")
                    )

Heatmap(mat, 
        show_column_names = F, 
        col=cols, 
        top_annotation = ha, 
        column_split = seurat_analysed@meta.data$SingleR2, 
        border = T, 
        cluster_column_slices = FALSE,
        cluster_row_slices =  FALSE,
        border_gp = gpar(col = "darkgrey", lty = 1),
        row_gap = unit(1.5, "mm"),
        row_split = top5$cluster,
        row_title = "DE Genes",
        column_title_side = "bottom", 
        column_title_rot = 90, 
        name="Expression", 
        left_annotation = ra
)
```

## 3.4 - Write to file
The annotation is done, the file is save on **_`r params$dex_save_file`_**.

```{r, include=FALSE}
### All the QC is done, so now we may write to file
write_rds(seurat_analysed, file=params$dex_save_file)
```
